Full Dissertation

### 2024-04-14 Sections/HTW.qmd
finish up motivation portion of intro

To fit ALM and EXAM to our participant data, we employ a similar method to @mcdanielPredictingTransferPerformance2009, wherein we examine the performance of each model after being fit to various subsets of the data. Each model was fit to the data with three separate procedures: 1) fit to maximize predictions of the testing data, 2) fit to maximize predictions of both the training and testing data, 3) fit to maximize predictions of the just the training data. We refer to this fitting manipulations as "Fit Method" in the tables and figures below. It should be emphasized that for all three fit methods, the ALM and EXAM models behave identically - with weights updating only during the training phase.Models were fit separately to the data of each individual participant. The free parameters for both models are the generalization ($c$) and learning rate ($lr$) parameters. Parameter estimation was performed using approximate bayesian computation (ABC), which we describe in detail below.

>>>>

To fit ALM and EXAM to our participant data, we employ a similar method to @mcdanielPredictingTransferPerformance2009, wherein we examine the performance of each model after being fit to various subsets of the data. Each model was fit to the data with three separate procedures: 1) fit to maximize predictions of the testing data, 2) fit to maximize predictions of both the training and testing data, 3) fit to maximize predictions of the just the training data. We refer to this fitting manipulations as "Fit Method" in the tables and figures below. It should be emphasized that for all three fit methods, the ALM and EXAM models behave identically - with weights updating only during the training phase. Models were fit separately to the data of each individual participant. The free parameters for both models are the generalization ($c$) and learning rate ($lr$) parameters. Parameter estimation was performed using approximate bayesian computation (ABC), which we describe in detail below.

### 2024-04-14 Sections/IGAS.qmd
finish up motivation portion of intro

---
title: IGAS Project
subtitle: An instance-based model account of the benefits of varied practice in visuomotor skill
date: last-modified
#page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
format:
  html: default
  hugo-md:
    echo: false
    include: false
    html-math-method: mathjax
    output-file: igas_hugo.md
execute: 
  warning: false
  eval: true
editor: 
  markdown: 
    wrap: 72
---

@@

The present work examines whether the commonly observed benefits of varied training can be accounted for by between-group differences in similarity between training and testing throws. We first attempt to replicate previous work finding an advantage of varied training over constant training in a projectile launching task. We then examine the extent to which this advantage can be explained by an instance-based similarity model.

>>>>

---
title: IGAS Project
subtitle: An instance-based model account of the benefits of varied practice in visuomotor skill
#date: last-modified
#page-layout: full
code-fold: true
code-tools: true
toc: true
toc-location: body
toc-title: Contents
format:
  html: default
  hugo-md:
    echo: false
    include: false
    html-math-method: mathjax
    output-file: igas_hugo.md
execute: 
  warning: false
  eval: true
editor: 
  markdown: 
    wrap: 72
---

@@

The present work examines whether the commonly observed benefits of varied training can be accounted for by a theoretrically motivated measure of the similarity between training throws and the testing solution space. We first attempt to replicate previous work finding an advantage of varied training over constant training in a projectile launching task. We then examine the extent to which this advantage can be explained by an instance-based similarity model.

### 2024-04-14 Sections/Introduction.qmd
finish up motivation portion of intro

>>>>

### Existing Theoretircal Frameworks

A number of theoretical frameworks have been proposed to conceptually explain the effects of varied training on learning and generalization. Schema theory (described in more detail above), posts that varied practice leads to the formation of more flexible motor schemas, which then facilitate generalization [@schmidtSchemaTheoryDiscrete1975]. The desirable difficulties framework [@bjorkMakingThingsHard2011; @soderstromLearningPerformanceIntegrative2015] proposes that variable practice may impair initial performance but then enhance longer-term retention and transfer. Similarily, the challenge point framework [@guadagnoliChallengePointFramework2004] contends that training variation induces optimal learning occurs insofar as it causes the difficulty of practice tasks to be appropriately matched to the learner's capabilities, but may also be detrimental if the amount of variation causes the task to be too difficult.

While these frameworks offer valuable conceptual accounts, there has been a limited application of computational modeling efforts aimed at quantitatively assessing and comparing the learning and generalization mechanisms which may be underlying the influence of variability in visuomotor skill learning. In contrast, the effects of variability have received more formal computational treatment in other domains, such as category learning [@hahnEffectsCategoryDiversity2005 @huHighvariabilityTrainingDoes2024],  language learning [@jonesDensityDistinctivenessEarly2020], and function learning [@deloshExtrapolationSineQua1997]. A primary goal of the  current dissertation is to to address this gap by adapting and applying modeling approaches from these other domains to investigate the effects of training variability in visuomotor skill learning and function learning tasks.

### 2024-04-14 Sections/hold_temp.qmd
finish up motivation portion of intro

>>>>

___________
___________

A number of theoretical frameworks have been proposed to conceptually explain the effects of varied training on learning and generalization, such as the contextual interference effect (Battig, 1972; Shea & Morgan, 1979), the desirable difficulties framework (Bjork, 1994; Soderstrom & Bjork, 2015), and the challenge point hypothesis (Guadagnoli & Lee, 2004). While these frameworks offer conceptual accounts of how variability influences skill acquisition and transfer, there has been a relative dearth of computational modeling efforts aimed at precisely specifying the underlying mechanisms, particularly within the domain of visuomotor skill learning.

In contrast, the effects of variability have received more formal computational treatment in other domains. For instance, in category learning, researchers have successfully applied exemplar models (e.g., Nosofsky, 1986) and connectionist architectures to capture how the diversity of training instances impacts generalization (Hahn et al., 2005; Hu & Nosofsky, 2024). Similarly, neural network models have been used to elucidate the influence of variability on learning and generalization in language acquisition (Jones & Brandt, 2020). However, such computational approaches have seen limited application to understanding the role of varied practice in visuomotor skill learning and transfer.

Given the practical relevance of optimizing training for generalization across domains, there is a clear need for formal computational models that can provide a precise, quantitative account of the cognitive processes through which varied practice impacts the acquisition and flexible expression of visuomotor skills. Addressing this gap would not only advance theoretical understanding but could also inform the design of more effective training regimens.

___________

Several influential theoretical frameworks, such as the contextual interference effect (Battig, 1972; Shea & Morgan, 1979), the desirable difficulties framework (Bjork, 1994; Soderstrom & Bjork, 2015), and the challenge point hypothesis (Guadagnoli & Lee, 2004), have been proposed to provide conceptual explanations for the effects of training variability on learning and generalization in sensorimotor skill acquisition. These frameworks suggest that introducing appropriate levels of variability or difficulty during training can enhance long-term retention and transfer of skills.

However, while these frameworks offer valuable conceptual accounts, there is currently a lack of formal computational modeling approaches that explicitly capture the mechanisms underlying the effects of varied training in the sensorimotor learning domain. In contrast, computational models have been developed to account for the influence of variability in other domains, such as category learning (Hahn et al., 2005; Hu & Nosofsky, 2024) and language learning (Jones & Brandt, 2020). For example, Hahn et al. (2005) used the generalized context model, an exemplar-based model, to explain how category variability affects generalization in a perceptual categorization task. Similarly, Jones and Brandt (2020) employed neural network simulations to investigate the role of density and distinctiveness in early word learning.

Given the importance of understanding the cognitive processes involved in sensorimotor learning and generalization, there is a need for computational models that can provide a more precise and quantitative account of how varied practice influences the acquisition and generalization of sensorimotor skills. Such models would complement the existing theoretical frameworks by specifying the underlying mechanisms and parameters that give rise to the observed effects of training variability, generating testable predictions, and offering a deeper understanding of the cognitive processes involved in sensorimotor learning and transfer.

___________

Several influential theoretical frameworks have been proposed to account for the effects of training variability on learning and generalization. The contextual interference effect (Battig, 1972; Shea & Morgan, 1979) posits that varied practice leads to greater interference during acquisition but enhanced retention and transfer due to increased contextual discrimination and cognitive effort during learning. Similarly, the reconceptualized desirable difficulties framework (Bjork, 1994; Soderstrom & Bjork, 2015) suggests that varied practice introduces desirable difficulties that initially impair performance but promote long-term retention and transfer. The challenge point hypothesis (Guadagnoli & Lee, 2004) proposes that optimal learning occurs when the difficulty of the task is matched to the learner's skill level, with varied practice serving to increase task difficulty and facilitate learning for more skilled individuals.

While these frameworks offer conceptual explanations for the influence of variability, the sensorimotor skill learning literature has been relatively limited in its application of formal computational modeling approaches to elucidate the underlying cognitive processes. In contrast, researchers have made significant progress in developing computational models to account for the effects of variability in other domains, such as category learning (e.g., Hahn et al., 2005; Nosofsky, 1986) and language learning (e.g., Jones & Brandt, 2020). The present work seeks to address this gap by implementing and evaluating computational models that can provide a more precise and quantitative account of how varied practice influences the acquisition and generalization of sensorimotor skills.

___________

While several influential frameworks provide conceptual insights into the role of training variability in sensorimotor skill learning, there remains a notable gap in the formal computational modeling of these effects. This contrasts sharply with the fields of category learning and language learning, where computational models have been extensively developed and applied. For instance, in the realm of category learning, Hahn, Bailey, and Elvin (2005) demonstrate how category diversity influences learning, memory, and generalization through computational modeling approaches that quantify the effects of category structure and exemplar variability. Similarly, in language learning, Jones and Brandt (2020) employed neural network simulations to explore how phonological neighborhood density and distinctiveness impact early word learning, providing a computational account of the complex dynamics involved.

In contrast, the sensorimotor domain has largely relied on qualitative frameworks such as the contextual interference effect and the challenge point framework without a corresponding development in computational models that can rigorously test these theories. For example, although studies like that of Hu and Nosofsky (2024) investigate the effects of high-variability training in the prototype-distortion paradigm, they primarily focus on empirical evaluations without integrating these findings into a broader computational framework. This discrepancy highlights a crucial need for developing computational models that can offer a more precise and mechanistic understanding of how variability affects the acquisition and generalization of sensorimotor skills, akin to the advances seen in category and language learning studies. Such efforts would not only bridge the theoretical gaps but also enhance the predictive power and applicability of these models in practical training and rehabilitation settings.

___________

A number of theoretical frameworks have been proposed to conceptually explain the effects of varied training on learning and generalization. Schema theory (described in more detail above), posts that varied practice leads to the formation of more flexible motor schemas, which then facilitate generalization [@schmidtSchemaTheoryDiscrete1975]. The desirable difficulties framework [@bjorkMakingThingsHard2011; @soderstromLearningPerformanceIntegrative2015] proposes that variable practice may impair initial performance but then enhance longer-term retention and transfer. Similarily, the challenge point framework [@guadagnoliChallengePointFramework2004] contends that training variation induces optimal learning occurs insofar as it causes the difficulty of practice tasks to be appropriately matched to the learner's capabilities, but may also be detrimental if the amount of variation causes the task to be too difficult.

While these frameworks offer valuable conceptual accounts, there has been a limited application of computational modeling efforts aimed at quantitatively assessing and comparing the learning and generalization mechanisms which may be underlying the influence of variability in visuomotor skill learning. In contrast, the effects of variability have received more formal computational treatment in other domains, such as category learning [@hahnEffectsCategoryDiversity2005 @huHighvariabilityTrainingDoes2024],  language learning [@jonesDensityDistinctivenessEarly2020], and function learning [@deloshExtrapolationSineQua1997]. A primary goal of the  current dissertation is to to address this gap by adapting and applying modeling approaches from these other domains to investigate the effects of training variability in visuomotor skill learning and function learning tasks.
